---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nadaverse: Browse Microdata Catalogs Using NADA REST API

<!-- badges: start -->
[![R-CMD-check](https://github.com/GutUrago/nadaverse/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/GutUrago/nadaverse/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/GutUrago/nadaverse/graph/badge.svg)](https://app.codecov.io/gh/GutUrago/nadaverse)
<!-- badges: end -->

`nadaverse` is the essential R package for researchers, policy analysts, and data enthusiasts seeking streamlined, programmatic access to vast collections of global microdata.

Many national and international organizations—including the World Bank, IHSN, FAO, UNHCR, and ILO—use the National Data Archive (NADA) software to manage and disseminate their survey and census data. While these catalogs are rich sources of information, interacting with them often requires tedious manual browsing or complex API construction.

`nadaverse` cuts through that complexity. It provides a unified, reliable, and user-friendly interface to search, filter, and retrieve crucial metadata and documentation (such as file lists and data dictionaries) directly into your R environment.


## Installation

You can install the stable version of this package by running:

``` r
install.packages("nadaverse")
```

or development version like so:

``` r
devtools::install_github("guturago/nadaverse")
```


## Searching

```{r}
library(nadaverse)
search_catalog("fao")

```



## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
#library(nadaverse)
# x <- search_catalog("unhcr", "FIES", ps = 10000) |> 
#   dplyr::filter(grepl("Food Insecurity Experience Scale", title, TRUE)) |> 
#   dplyr::select(nation, year_start) |> 
#   dplyr::arrange(nation, year_start)
# 
# y <- mutate(x,
#             value = "Yes") |> 
#   pivot_wider(id_cols = nation, 
#               names_from = year_start, 
#               values_from = value,
#               values_fill = "-") |> 
#   gt::gt()
# y
```

